Reinforcement Learning Methods for Operations Research Applications -- Reloaded

### 2020-12-16 paper/paper.tex
minor updates paper

\begin{abstract}
  % teilw. von IJPR paper uebernommen
  An essential task in manufacturing planning and control is to determine when to release orders to
  the shop floor. One key parameter is the lead time which is the planned time that elapses between
  the release of an order and its completion. Lead times are normally determined based on the actual
  duration orders previously took to traverse through the production system (flow times).
  Traditional order release models assume static lead times, although it has been shown that they
  should be set dynamically to react the dynamic operational characteristics of the system.
  Therefore, we present an order release model which sets lead times dynamically by using an
  reinforcement learning approach. Therefore, we provide an in-depth analysis of reinforcement
  learning to show that average reward reinforcement learning is a better approach for operations
  research applications as discounted reinforcement learning. Additionally we present an average
  reward reinforcement learning algorithm which infers
  (Blackwell-)optimal\footnote{Blackwell-optimality is not yet fully proven! A first version of such
    a proof is attached in the Appendix.} policies.
  % \SH{Add RL stuff}
  % Therefore, we present a flow time estimation procedure to set lead times dynamically
  % by using an artificial neural network. Additionally, we implement a safety lead time
  % to incorporate the underlying cost ratio in the order release model. We test our
  % proposed approach using a simulation model of a three-stage make-to-order flow-
  % shop and compare the forecast accuracy and the accumulated costs to other forecast-
  % based order release models from literature. We show that our proposed model using
  We use a simulation model of a two-stage flow-shop to compare the algorithm
  to % and show that our approach outperforms
  well-known order release mechanisms. We show that in the current version our proposed model using
  reinforcement learning outperforms some, but not all other tested approaches.
  % EVTL: especially for scenarios with...
  \keywords{operations research, production planning, order release, machine learning, reinforcement
    learning}
\end{abstract}

\section{Introduction}
\label{sec:introduction} An important goal in Manufacturing Planning and Control (MPC) systems is to
achieve short and predictable flow times, especially where high flexibility in meeting customer
demand is required. Besides achieving short flow times, one should also maintain high output and
due-date performance while keeping the work-in-process level low. One approach to address this
problem is to collect all incoming orders in an order-pool and periodically decide which orders to
release to the shop floor. Once orders are released, costs start to accumulate as planned orders
materialise as actual jobs in the production system. The main challenge is to find a good compromise
of balancing the shop floor and timely completion of jobs. Although the performance of such systems
can be measured manifold the most overarching objective is to maximise profits by adequately
assigning holding and lateness costs.

One of the key modeling parameters for order release mechanisms is the \textit{lead time} which
refers to the planned time that elapses between the release of an order and its arrival in the
finished goods inventory. This planning parameter is often based on the observable time an order
needs to traverse through the production system, which we denote as \textit{flow time}. Flow times
consist of processing, setup, control, transport, and waiting times, whereas the latter is the
governing factor. Waiting times are a result from queuing (e.g. jobs queue before and after
processing) which depend heavily on the amount of jobs in the system (WIP) and thus, waiting times
are relatively difficult to estimate which makes the setting of favorable lead times so difficult
(e.g., \citealt{Tatsiopoulos1983, Wiendahl1995}). Most state-of-the-art order release mechanisms use
static or fixed lead times to address the order release problem, and thus neglect the nonlinear
relationship between resource utilisation and flow times which is well known from practice and
queuing theory (\citealt{Pahl2007}).

One way to address this nonlinear interaction effects is to set lead times dynamically. Thus,
intuitively the order release problem is solved by perfectly matching the lead times to the flow
times, but one faces ``sampling issues'' meaning that the flow times depend on the lead times. An
extreme scenario of this problem is the so called ``lead time syndrome'' which describes a vicious
cycle where increasing flow times perpetually inflate the lead times which leads to worse
performance (see e.g., \citealt{Mather1978, knollmann2013control, Selcuk2006}. Therefore,
\cite{SchneckenreitherHaeussler2019} propose to use reinforcement learning as approach to
incorporate these effects in this optimisation problem. Reinforcement learning is an optimisation
technique that stems from dynamic programming. The goal in reinforcement learning is to find the
best stationary policy for a given problem. This policy is usually provided by assessing the states
(or state-action pairs) of an underlying Markov Decision Process (MDP). The advantage of
reinforcement learning over dynamic programming is that (i) the problem space is explored by an
agent and thus only expectantly interesting parts of the problem space need to be assessed and (ii)
the knowledge (acquisition) of transition probabilities becomes unnecessary as the states are
evaluated by consecutively observed states solely.

Over the past decades reinforcement learning has been applied to various problems, for which
astonishing results have been reported. E.g. only recently \cite{mnih2015human} have presented a novel
value-iteration reinforcement learning agent which exceeds human-level abilities in playing many
classic Atari 2600 games~\citep{bellemare2012investigating}. \cite{mnih2016asynchronous} presents
improved results with asynchronous actor-critic reinforcement learning. Also games like
Go~\citep{silver2016mastering} and Chess~\citep{silver2017mastering} have been mastered with
superhuman performance by \textit{tabula rasa} reinforcement learning agents. Furthermore, the
method has also been applied in the setting of manufacturing system, e.g. to improve the ramp-up
process~\citep{doltsinis2012reinforcement}, in locally selecting appropriate dispatching rules
\citep{zhang1995reinforcement,wang2005application} or scheduling
\citep{zhang1995reinforcement, waschneck2018optimization}.
%
However, all these applications use discounted reinforcement learning and are either designed to
investigate a rather simple MDP, e.g. by selecting heuristics instead of optimising the underlying
problem itself, or by mapping the actual objective in a reward function that is approximately \(0\)
on average over time.
%
This is due to the fact that state value is largely composed of a term defined by the policy's
average reward value \citep{MillerVeinott1969,Blackwell62} which would otherwise dilute the state
values and thus decrease the solution quality, as for instance in
\cite{SchneckenreitherHaeussler2019} and \cite{gijsbrechts2018can}.
%

Therefore, most applications that incorporate and directly reflect costs or profit in the reward
function use average reward reinforcement learning. For instance \cite{aydin2000dynamic} use it in
the setting of scheduling, while in a series of papers Mahadevan et al. investigated several problem
domains starting with simple MDPs
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
After these foundational works for average reward reinforcement learning they introduced a
continuous time average reward reinforcement learning algorithm named SMART
\citep{Mahadevan97_SelfimprovingFactorySimulationUsingContinuoustimeAveragerewardReinforcementLearning}.
Applications of SMART reach from the optimisation of queuing systems
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies,Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning},
maintenance of an inventory system
\citep{Das99_SolvingSemiMarkovDecisionProblemsUsingAverageRewardReinforcementLearning} to optimising
transfer line in terms of maximizing demand, while keeping inventory levels of unfinished product as
low as possible \citep{Mahadevan98_OptimizingProductionManufacturingUsingReinforcementLearning}.

Like discounted reinforcement learning also average reward reinforcement learning is based on an
oracle function, in our case the accumulated profit of a period, to assess the decisions taken by
the agent. By repeatedly choosing different actions the agent examines the problem space and rates
possible actions for any observed state. The advantage of average reward reinforcement learning over
the widely applied discounted reinforcement learning framework is that the underlying optimisation
technique is able to find better policies. This yields from the fact that in the latter method the
states are assessed by a single value, while in the former the values are split up by the average
reward accumulated over time and a bias value which optimises the number of steps to reach the best
possible states of the examined process
\citep{Howard64,Puterman94,Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.
Thus, while in average reward reinforcement learning these terms are learned separately in the
discounted framework one value consisting of the addition of these values plus an additional error
term is estimated. This incautious combining of different kinds of state values as done in
discounted reinforcement learning leads to the problems that (i) the average reward, which is equal
for all states of usually investigated unichain MDPs, is dominating and thus diluting the bias
values, and (ii) the state values are deteriorated by the error term that is only imposed due to the
discounting technique.

Thus, as opposed to the commonly applied discounted reinforcement learning algorithm and as proposed
as future work by \cite{SchneckenreitherHaeussler2019} we use an average reward reinforcement
learning algorithm to adaptively release orders based on the assessed state values of the production
system. In contrast to the aforementioned works on average reward reinforcement learning our
algorithm incorporates the optimisation of not only the average reward over time, but also the bias
values. Only the work by
\cite{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}
integrates this second-level refinement optimisation. However, their algorithm requires the
selection of a reference state to prevent an infinite increase of state values. This results from
the lack of feedback in the iterative process of optimising the different refinement optimisation
levels.

@@

\paragraph{Structure.} The rest of the paper is structured as follows. The next section introduced
average reward reinforcement learning and presents the used algorithm (see also
\citealt{Schneckenreither2019}) including a proof-of-concept.
Section~\ref{sec:Experimental_Evaluation} describes the simulation model we use to evaluate the
approach. Then~\ref{sec:Preliminary} provides preliminary result data, wheras
Section~\ref{sec:conclusion} concludes this working paper.

\section{Average Reward Reinforcement Learning}
\label{sec:Average_Reward_Reinforcement_Learning}

This section introduces average reward reinforcement learning in comparison to the discounted
framework, elaborates on optimiality criteria and provides insights of the underlying algorithm.

@@

In the discounted framework the value of a state \(V_{\gamma}^{\pol}(s)\) is defined as the expected
discounted sum of rewards under the stationary policy \(\pol\) when starting in state \(s\). That is
\begin{align*}
  V_{\gamma}^{\pol}(s)=\lim_{N \to \infty} E[\sum_{t=0}^{N-1} \gamma^{t} R_{t}^{\pol}(s)]\tcom
\end{align*}

@@

The aim in the discounted framework is to find an optimal policy \(\polopt\), which when followed,
maximises the state value for all states \(s\) as compared to any other policy \(\pol\):
\(V_{\gamma}^{\polopt} - V_{\gamma}^{\pol} \geqslant 0\). This criteria is usually referred to as
\(\gamma\)-optimality as the discount factor \(\gamma\) is fixed. However, this also means that the
actual value set for \(\gamma\) defines the policy that is optimised for. For instance, as can be
seen in Figure~\ref{fig:printer} setting \(\gamma\) value to \(<0.8027\) the printer-loop is
preferred over the mail-loop by the agent, although the mail-loop accumulates more reward over time,
i.e. selecting the mail-loop is a sub-optimal choice.

@@

  \end{tikzpicture}
  \caption{\label{fig:printer} A MDP with two different determinstic policies. The only action choice is in
    state \(1\), in which the agent can choose between doing the printer-loop or the mail-loop.
    Observe that the average reward received per step equals \(1\) for the printer-loop and \(2\)
    for the mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However,
    if \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement
    learning chooses the printer loop.
    %
    Adapted from \cite{Mahadevan96_OptimalityCriteriaInReinforcementLearning}.}
\end{figure}

The main idea behind the \(\gamma\)-parameter of the discounted framework is to be able to balance
short-term rewards (low \(\gamma\)-values) and long-term rewards (high \(\gamma\)-values). However,
what seems to be an advantage rather becomes a disadvantage for most applications. The issue is that
the average reward value is non-linearly increased when \(\gamma\) approaches one. However, in
almost all cases the aim is to perform well over time which in terms of reward means to first
maximise for a the policy with highest average reward before more selectively choosing actions. That
is, if several processes exist, we are searching for the ones with highest average reward before
considering other criteria.
%
Therefore, in almost all RL studies the discount factor is set to a value very close to \(1\), for
instance to \(0.99\)~\cite[e.g.]{mnih2015human,mnih2016asynchronous,Lillicrap15}.
%
By doing so the above mentioned non-linear relationship leads to the fact that the state value
consists almost solely of the up-scaled average reward term, whereas the bias values can be
neglected. This leads to diluted state values and thus difficulties in distinguishing actions which
impose policies that possess the same average reward.
%
In contrast to that, average reward RL separately assesses these values and according to these
selectively chooses the best action, comparing one after the other.
%
To illustrate the idea reconsider Figure~\ref{fig:printer}. First the agent picks the policy
according to the highest average reward. Thus, all actions are assessed by an average reward of
\(2\) which is inferred from the Mail-loop. This however, makes choosing the Printer-loop
unattractive; i.e. the policy converges to the optimal policy of choosing the Mail loop.

\subsection{Average Reward Reinforcement Learning}
\label{subsec:Average_Reward_Reinforcement_Learning}

@@

The addends, where \(n=-1,0,\ldots\) denote the coefficients of the Laurent series expansion,
can be reformulated to following constraint problem with expected reward
\(R^{\pol}(s) = E_{\pol}[ r(s, a) ] = \sum_{a \in \mathcal{A}(s)}p(a | s) r(s, a)\) as shown by
\cite{MillerVeingsott1969} and \citet[p.346]{Puterman94}:

@@

\citet[p.343ff]{Puterman94} shows that due to the given degree of freedom if \(n=-1,0,\ldots,M\)
constraints are satisfying the above conditions for all states \(s\), then only
\(\avgrew^{\pol}(s), V^{\pol}(s), W^{\pol}_{1}, \ldots, W^{\pol}_{M-1}\) are unique, whereas
\(W^{\pol}_{M}\) is offset by the vector \(u\) where for the transition probability matrix \(P\) the
vector \(u\) is characterised by \((I-P)u = 0\). Note that \(u\) is determined by the number of
closed irreducible classes of \(P\), that is for ergodic MDPs \(u\) is determined by a single
constant. Our algorithm is based on the above formulation.

@@

This part proves (or rather ought to prove, as this is only the first version of such a proof)
Blackwell-optimality of the average reward reinforcement learning algorithm shown in
Algorithm~\ref{alg}. The idea is to prove \(n\)-discount optimality for \(n=-1\), \(n=0\), and
finally for \(n \geqslant 1\). But first recall the definition of \(n\)-discount-optimality.

>>>>

\begin{abstract}
  % teilw. von IJPR paper uebernommen
  An essential task in manufacturing planning and control is to determine when to release orders to
  the shop floor. One key parameter is the lead time which is the planned time that elapses between
  the release of an order and its completion. Lead times are normally determined based on the actual
  duration orders previously took to traverse through the production system (flow times).
  Traditional order release models assume static lead times, although it has been shown that they
  should be set dynamically to react the dynamic operational characteristics of the system.
  Therefore, we present an order release model which sets lead times dynamically by using an
  reinforcement learning approach. Therefore, we provide an in-depth analysis of reinforcement
  learning to show that average reward reinforcement learning is a better approach for operations
  research applications as discounted reinforcement learning. Additionally we present an average
  reward reinforcement learning algorithm which infers near-Blackwell-optimal policies. We use a
  simulation model of a \MS[t]{todo}two-stage flow-shop to compare the algorithm
  to % and show that our approach outperforms
  well-known order release mechanisms. We show that in the current version our proposed model using
  reinforcement learning outperforms some, but not all other tested approaches.
  % EVTL: especially for scenarios with...
  \keywords{operations research, production planning, order release, machine learning, reinforcement
    learning}
\end{abstract}

\section{Introduction}\label{sec:introduction}
An important goal in Manufacturing Planning and Control (MPC) systems is to achieve short and
predictable flow times, especially where high flexibility in meeting customer demand is required.
Besides achieving short flow times, one should also maintain high output and due-date performance
while keeping the work-in-process level low. One approach to address this problem is to collect all
incoming orders in an order-pool and periodically decide which orders to release to the shop floor.
Once orders are released, costs start to accumulate as planned orders materialise as actual jobs in
the production system. The main challenge is to find a good compromise of balancing the shop floor
and timely completion of jobs. Although the performance of such systems can be measured manifold the
most overarching objective is to maximise profits by adequately assigning holding and lateness
costs.

One of the key modeling parameters for order release mechanisms is the \textit{lead time}, which
refers to the planned time that elapses between the release of an order and its arrival in the
finished goods inventory. This planning parameter is often based on the observable time an order
needs to traverse the production system, which in contrast is denoted as the \textit{flow time}.
Flow times consist of processing, setup, control, transport, and waiting times, whereas the latter
is the governing factor \citep[][p.223]{zapfel1982produktionswirtschaft}. Waiting times are a result
from queuing (e.g.\@ jobs queue before and after processing), depend heavily on the amount of jobs in
the system (WIP) and thus are relatively difficult to estimate, which makes the
setting of favorable lead times so difficult (e.g., \citealt{Tatsiopoulos1983, Wiendahl1995}). Most
state-of-the-art order release mechanisms use static or fixed lead times to address the order
release problem, and thus neglect the nonlinear relationship between resource utilisation and flow
times, which is well known from practice and queuing theory (\citealt{Pahl2007}).

One way to address this nonlinear interaction effects is to set lead times dynamically. Intuitively
the order release problem is solved by perfectly matching the lead times to the flow times, but the
corresponding optimisation problem faces ``sampling issues'' meaning that the flow times depend on
the lead times. An extreme scenario of this problem is the so called ``lead time syndrome'', which
describes a vicious cycle where increasing flow times perpetually inflate the lead times which leads
to worse performance (see e.g., \citealt{Mather1978, knollmann2013control, Selcuk2006}. Therefore,
\citet{SchneckenreitherHaeussler2019} propose to use reinforcement learning as approach to
incorporate these effects in this optimisation problem. Reinforcement learning is an optimisation
technique that stems from dynamic programming and its goal is to find the best stationary policy for
a given problem. This policy is usually provided by assessing the states (or state-action pairs) of
an underlying Markov Decision Process (MDP). The advantage of reinforcement learning over dynamic
programming is that (i) the problem space is explored by an agent and thus only expectantly
interesting parts of the problem space need to be assessed and (ii) the knowledge (acquisition) of
transition probabilities becomes unnecessary as the states are evaluated by consecutively observed
states solely.

Over the past decades reinforcement learning has been applied to various problems, for which
astonishing results have been reported. E.g.\@ only recently \citet{mnih2015human} have presented a
novel value-iteration reinforcement learning agent which exceeds human-level abilities in playing
many classic Atari 2600 games~\citep{bellemare2012investigating}. Further,
\citet{mnih2016asynchronous} present improved results with asynchronous actor-critic reinforcement
learning. Also games like Go~\citep{silver2016mastering} and Chess~\citep{silver2017mastering} have
been mastered with superhuman performance by \textit{tabula rasa} reinforcement learning agents.
Furthermore, the method has also been applied in the setting of manufacturing system, e.g.\@ to
improve the ramp-up process~\citep{doltsinis2012reinforcement}, in locally selecting appropriate
dispatching rules \citep{zhang1995reinforcement,wang2005application} or scheduling
\citep{zhang1995reinforcement, waschneck2018optimization}.
%
However, all these applications use discounted reinforcement learning and are either designed to
investigate a rather simple MDP, e.g.\@ by selecting heuristics instead of optimising the underlying
problem itself, or by mapping the actual objective in a reward function that is approximately \(0\)
on average over time.
%
This is due to the fact that state value is largely composed of a term defined by the policys'
average reward value \citep{MillerVeinott1969,Blackwell62} which would otherwise dilute the state
values and thus decrease the solution quality, as can for instance be observed in
\citet{SchneckenreitherHaeussler2019} and \citet{gijsbrechts2018can}.
%

Therefore, most applications that incorporate and directly reflect costs or profit in the reward
function use average reward reinforcement learning. \citet{aydin2000dynamic} use it in the setting
of scheduling, while in a series of papers Mahadevan et al. investigated several problem domains
starting with simple MDPs
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
After these foundational works for average reward reinforcement learning they introduced a
continuous time average reward reinforcement learning algorithm named SMART
\citep{Mahadevan97_SelfimprovingFactorySimulationUsingContinuoustimeAveragerewardReinforcementLearning}.
Applications of SMART reach from the optimisation of queuing systems
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies,Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning},
maintenance of an inventory system
\citep{Das99_SolvingSemiMarkovDecisionProblemsUsingAverageRewardReinforcementLearning} to optimising
transfer line in terms of maximizing demand, while keeping inventory levels of unfinished product as
low as possible \citep{Mahadevan98_OptimizingProductionManufacturingUsingReinforcementLearning}.
However, in practise usually decision have to be made on a daily basis \citep{enns2004work}.
Therefore, we refrain from this adaption and concentrate on standard MDPs only. Furthermore, often
continuous-time semi-MDP problems can be converted through uniformisation into equivalent discrete
time instances \citep[see][]{Puterman94,bertsekas1995dynamic}.

Like discounted reinforcement learning also average reward reinforcement learning is based on an
oracle function, in our case the accumulated costs of a period, to assess the decisions taken by the
agent. By repeatedly choosing different actions the agent examines the problem space and rates
possible actions for any observed state. The advantage of average reward reinforcement learning over
the widely applied discounted reinforcement learning framework is that the underlying optimisation
technique is able to find better policies. This yields from the fact that in standard discounted
reinforcement learning method the states are assessed independently and by a single value. In
contrast to that average reward reinforcement learning splits up the evaluation of the average
reward per step, a bias value that specifies the amount of collected rewards to reach the optimal
path when starting in a suboptimal state and an error term which defines the number of steps to
reach the optimal path
\citep{Howard64,Puterman94,Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.
Thus, while in average reward reinforcement learning these terms are learned separately in the
discounted framework one value consisting of the addition of these values is estimated, where
however the average reward is scaled by \(1/(1-\gamma)\) with the discount factor \(\gamma\) usually
being set very close to \(1\), e.g. \(0.99\). In commonly applied unichain MDPs the average reward
per step is equal for all states. Thus, independently assessing it is computationally unwisely.
Therefore, we adapt an algorithm that uses a scalar value for the estimation of the average reward.
Furthermore, as the average reward is scaled in the discounted state values it dominates the other
two terms. This, combined with the iterative evaluation and the independently assessing of the
average reward for each state lets standard discounted reinforcement learning struggle to find good
policies.
% Thus, this incautious combining of different kinds of state values as done in discounted
% reinforcement learning leads to the problems that (i) the average reward, which is equal for all
% states of usually investigated unichain MDPs, is dominating and thus diluting the bias values, and
% (ii) the state values are deteriorated by the error term that is only imposed due to the
% discounting technique.

Thus, as opposed to the commonly applied discounted reinforcement learning algorithm we use an
average reward adjusted reinforcement learning algorithm to adaptively release orders based on the
assessed state values of the production system. In contrast to the aforementioned works on average
reward reinforcement learning our algorithm incorporates the optimisation of not only the average
reward over time, but also the bias values, which is an important adaption in highly stochastic
systems. Only the work by
\cite{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}
integrates this second-level refinement optimisation. However, their algorithm requires the
selection of a reference state to prevent an infinite increase of state values. This results from
the lack of feedback in the iterative process of optimising the different refinement optimisation
levels. Furthermore, they asses the average reward independently.

@@

\paragraph{Structure.} The rest of the paper is structured as follows. The next section introduced
average reward reinforcement learning and presents the used algorithm (see also
\citealt{schneckenreither2020average}). Section~\ref{sec:Experimental_Evaluation} describes the
simulation model we use to evaluate the approach. \MS[t]{todo}
% Then~\ref{sec:Preliminary} presents preliminary
% result data, whereas Section~\ref{sec:conclusion} concludes this working paper.

\section{Average Reward Adjusted Reinforcement Learning}

This section introduces average reward adjusted reinforcement learning in comparison to the discounted
framework, elaborates on optimiality criteria and provides insights of the underlying algorithm.

@@

In the widely applied discounted framework the value of a state \(V_{\gamma}^{\pol_{\gamma}}(s)\) is
defined as the expected discounted sum of rewards under the stationary policy \(\pol_{\gamma}\) when
starting in state \(s\). Note that the policy \(\pol_{\gamma}\) depends on the selected discount
factor. That is
\begin{align*}
  V_{\gamma}^{\pol_{\gamma}}(s)=\lim_{N \to \infty} E[\sum_{t=0}^{N-1} \gamma^{t} R_{t}^{\pol_{\gamma}}(s)]\tcom
\end{align*}

@@

The aim in the discounted framework is to find an optimal policy \(\polopt_{\gamma}\), which when
followed, maximises the state value for all states \(s\) as compared to any other policy
\(\pol_{\gamma}\): \(V_{\gamma}^{\polopt_{\gamma}} - V_{\gamma}^{\pol_{\gamma}} \geqslant 0\). This
criteria is usually referred to as \(\gamma\)-optimality as the discount factor \(\gamma\) is fixed.
However, this also means that the actual value set for \(\gamma\) defines the policy that is
optimised for. For instance, as can be seen in Figure~\ref{fig:printer} setting \(\gamma\) value to
\(<0.8027\) the printer-loop is preferred over the mail-loop by the agent, although the mail-loop
accumulates more reward over time, i.e. selecting the mail-loop is a sub-optimal choice.

@@

  \end{tikzpicture}
  \caption{\label{fig:printer} A MDP with two different deterministic policies. The only action choice is in
    state \(1\), in which the agent can choose between doing the printer-loop or the mail-loop.
    Observe that the average reward received per step equals \(1\) for the printer-loop and \(2\)
    for the mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However,
    if \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement
    learning chooses the printer loop.
    %
    Adapted from \cite{Mahadevan96_OptimalityCriteriaInReinforcementLearning}.}
\end{figure}

One idea behind the \(\gamma\)-parameter of the discounted framework is to be able to balance
short-term rewards (low \(\gamma\)-values) and long-term rewards (high \(\gamma\)-values). However,
what seems to be an advantage rather becomes a disadvantage for most applications. The issue is that
the average reward value is non-linearly increased when \(\gamma\) approaches one. However, in
almost all cases the aim is to perform well over time which in terms of reward means to first
maximise for a the policy with highest average reward before more selectively choosing actions. That
is, if several processes exist, we are searching for the ones with highest average reward before
considering other criteria.
%
Therefore, in almost all RL studies the discount factor is set to a value very close to \(1\), for
instance to \(0.99\)~\cite[e.g.]{mnih2015human,mnih2016asynchronous,Lillicrap15}.
%
By doing so the above mentioned non-linear relationship leads to the fact that the state value
consists almost solely of the up-scaled average reward term, whereas the bias values can be
neglected. This leads to diluted state values and thus difficulties in distinguishing actions which
impose policies that possess the same average reward.
%
In contrast to that, average reward RL separately assesses these values and according to these
selectively chooses the best action, comparing one after the other.
%
To illustrate the idea reconsider Figure~\ref{fig:printer}. First the agent picks the policy
according to the highest average reward. Thus, all actions are assessed by an average reward of
\(2\) which is inferred from the Mail-loop. This however, makes choosing the Printer-loop
unattractive; i.e. the policy converges to the optimal policy of choosing the Mail loop.

\subsection{Average Reward Reinforcement Learning}

@@

The addends, where \(n=-1,0,\ldots\) denote the coefficients of the Laurent series expansion,
can be reformulated to following constraint problem with expected reward
\(R^{\pol}(s) = E_{\pol}[ r(s, a) ] = \sum_{a \in \mathcal{A}(s)}p(a | s) r(s, a)\) as shown by
\cite{MillerVeinott1969} and \citet[p.346]{Puterman94}:

@@

\citet[p.343ff]{Puterman94} shows that due to the given degree of freedom if \(n=-1,0,\ldots,M\)
constraints are satisfying the above conditions for all states \(s\), then only
\(\avgrew^{\pol}(s), V^{\pol}(s), W^{\pol}_{1}, \ldots, W^{\pol}_{M-1}\) are unique, whereas
\(W^{\pol}_{M}\) is offset by the vector \(u\) where for the transition probability matrix \(P\) the
vector \(u\) is characterised by \((I-P)u = 0\). Note that \(u\) is determined by the number of
closed irreducible classes of \(P\), that is for ergodic MDPs \(u\) is determined by a single
constant. Average reward learning is based on the above formulation.

A major problem occurring at average reward RL is that the bias values are not uniquely defined
without solving the first set of constraints defined by the error term addends
\citep[see][p.346]{Puterman94,Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning}.
% We could overcome this issue by simply requiring \(\gamma\) to be strictly less than \(1\).
%
Our algorithm, based on the tabular version of \citet{schneckenreither2020average}, does not require
the exact solution for \(V^{\pol}(s)\), but a solution which is offset suffices. Clearly this
observation reduces the required iteration steps tremendously as finding the exact solution,
especially for large discount factors, is tedious. Therefore, we allow to set \(\gamma = 1\), which
induces \(\X_{\gamma}^{\pol}(s) = V^{\pol}(s) + u\), where \(u\) is for unichain MDPs a scalar value
independent of \(s\), i.e.\@ equivalent for all states of the MDP~\cite[p.346]{Puterman94}.
%
%% Nonetheless, we require \(\gamma < 1\) to stabilise the process.
%
If we are interested in correct bias values, i.e. \(\gamma\) is sufficiently close but strictly less
than \(1\), our approach is a tremendous advantage over average reward RL as it reduces the number
of iterative learning steps by requiring only a single constraint per state plus one for the scalar
average reward value. That is, for an MDP with \(N\) states only one more constraint (\(N+1)\) has
to be solved in \ARA{} as compared to (at least) \(2N+1\) nested constraints for average reward RL.
Therefore, it is cheap to compute \(\X_{\gamma}^{\pol}(s)\), while it is rather expensive to find
the correct values of \(V^{\pol}(s)\) directly, especially in an iterative manner as RL is.

@@

This part proves (or rather ought to prove, as this is only the first version of such a proof)
Blackwell-optimality of the average reward adjusted reinforcement learning algorithm shown in
Algorithm~\ref{alg:near}. The idea is to prove \(n\)-discount optimality for \(n=-1\), \(n=0\), and
finally for \(n \geqslant 1\). But first recall the definition of \(n\)-discount-optimality.
